---
title: "Military spending of NATO, Ukraine, and Russia since 1990 compared"
subtitle: "An exploratorive data analysis"
author: "Thomas McKinney"
date: "June 10th 2023"
output:
  html_document:
    toc: true
    number_sections: true
    theme: paper
    df_print: paged
---

```{r, Setup, message=FALSE}
library(tidyverse)
library(janitor)
library(here)
```
# Version:
Last change: June 10th 2023

# Introduction
I have recently started, to learn R. This is very much a first attempt at a R Notebook, and work in progress. I'm happy for any tipps or criticism. :-)

# Loading and arranging the Data
## % of Government spending
```{r, ReadData, message=FALSE}
dft_government_spending <- read_csv("240608_WB_MilSpend_GovtSpen_NATO_RUS_UKR_1990_2022.csv")
dft_government_spending <- as_tibble(dft_government_spending) %>% clean_names()
```

## Min, Max, Mean of % of Government spending by Country
```{r, message=FALSE}
dft_sum_government_spending <- dft_government_spending %>% 
  select(!c(series_code, country_name, natojoin, nato2024)) %>% 
  pivot_longer(!country_code, names_to = "year", values_to = "spending", values_drop_na = TRUE) %>% 
  group_by(country_code) %>%
  summarise(
    mean = mean(spending),
    min = min(spending),
    max = max(spending)
  ) %>% 
  arrange(desc(mean)) %>%
  as_tibble()
```

## Min, Max, Mean of % of Government spending by Year
```{r, message=FALSE}
dft_sum_government_spending_PA <- dft_government_spending %>% 
  select(!c(series_code, country_name, country_code, natojoin, nato2024)) %>% 
  pivot_longer(
    cols = starts_with("x"),
    names_to = "Year",
    names_prefix = "x",
    values_to = "spending",
    values_drop_na = TRUE
  ) %>% 
  mutate(Year = as.integer(Year)) %>% 
  arrange(desc(Year)) %>% 
  group_by(Year) %>% 
  summarise(
    mean = mean(spending),
    min = min(spending),
    max = max(spending)
  ) %>%
  as_tibble()
```


```{r}
dft_GDP <- read_csv("240608_WB_MilSpend_GDP_NATO_RUS_UKR_1990_2022.csv")
dft_GDP <- as_tibble(dft_GDP) %>% clean_names()

dft_sum_GDP <- dft_GDP %>% 
  select(!c(series_code, country_name, natojoin, nato2024)) %>% 
  pivot_longer(!country_code, names_to = "year", values_to = "spending", values_drop_na = TRUE) %>% 
  group_by(country_code) %>%
  summarise(
    mean = mean(spending),
    min = min(spending),
    max = max(spending)
  ) %>% 
  arrange(desc(mean)) %>% 
  as_tibble()

```


# Overview over the Data
## % of Government spending
```{r, OverviewGovtSPending}
ggplot(dft_sum_government_spending)+
  geom_errorbar(
    mapping = aes(
      x=country_code, ymin=min, ymax=max
      ), 
    width = 1
    )+
  geom_point(
    mapping = aes(
      x = country_code,
      y = mean,
    )
  )+
  scale_x_discrete(
    labels=abbreviate
  )+
  labs(
    title="Summary of Military spending as % of Government spending by Country", 
    subtitle = "Mean, Min and Max for the years 1990 - 2022 where available."
  )

dft_sum_government_spending
```
### Summary by year
```{r}
ggplot(dft_sum_government_spending_PA)+
  geom_errorbar(
    mapping = aes(
      x=Year, ymin=min, ymax=max
      ), 
    width = 1
    )+
  geom_point(
    mapping = aes(
      x = Year,
      y = mean,
    )
  )+
  labs(
    title="Summary of Military spending as % of Government spending by Year", 
    subtitle = "Mean, Min and Max for the years 1990 - 2022 where available."
  )

dft_sum_government_spending_PA
```

## % of GDP
```{r, OverviewGDP}
ggplot(dft_sum_GDP)+
  geom_errorbar(
    mapping = aes(
      x=country_code, ymin=min, ymax=max
      ), 
    width = 1
    )+
  geom_point(
    mapping = aes(
      x = country_code,
      y = mean,
    )
  )+
  scale_x_discrete(
    labels=abbreviate
  )+
  labs(
    title="Summary of Military spending as % of Government spending by Country", 
    subtitle = "Mean, Min and Max for the years 1990 - 2022 where available."
  )

dft_sum_GDP
```
## Summary by year
```{r}
summary(dft_GDP)
```